Stellar multiplicity in \Gaia

### 2018-05-28 article/ms.tex
words

\newcommand\teff{T_{\rm eff}}
\newcommand\logg{\log{g}}
\newcommand\feh{[\rm{Fe}/\rm{H}]}
\newcommand\mh{[\rm{M}/\rm{H}]}

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\package}[1]{\texttt{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\Gaia}{\project{Gaia}}
\newcommand{\gaia}{\project{gaia}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}

@@

%\submitjournal{AAS Journals}

\shorttitle{Stellar multiplicity}
\shortauthors{Casey et al.?}

@@

%\author{Daniel Foreman-Mackey}
%\affiliation{Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

@@

\begin{abstract}
Most stars are in binary systems. 
The occurrence rate of binary star systems is critically important to many 
fields of astrophysics, but the effects of binaries are commonly ignored in 
astronomical data analysis applications.
Here we construct a model to probabilistically identify binary stars based on
excess radial velocity scatter, photometry, and astrometric excess noise.

We identify binaries \todo{within some orbital and SP limits} and show that 
the orbital semi-amplitude is well-approximated by the excess radial velocity
scatter for systems within these criteria.

\todo{We make a few interesting tests about binarity...}
\end{abstract}

@@

There are at least three ways that binary systems can be reliably
identified using from \Gaia data: through radial velocity variations, 
anomalous photometric colours, or excess astrometric noise. In this 
work we do not make use of astrometric noise because of the 
degree-of-freedom astrometry bug present in the second \Gaia\ data 
release \citep{Lindegren:2018}. We will describe the radial velocity
information that forms part of our model for identifying binary stars,
and then discuss the photometric information used, before explicitly
outlining the assumptions of our model.

The second \Gaia\ data release provides median radial velocity 
estimates for 7,224,631 stars. The radial velocity error provided
($\GaiaRVE$; \texttt{radial\_velocity\_error}) is a function of the 
number of transits $N$ (i.e., observations; column 
\texttt{rv\_nb\_transits}) and the scatter among those $N$ measurements. 
This allows us to calculate the scatter among those measurements, or
rather the \emph{radial velocity jitter}, for each source
\begin{eqnarray}
\RVJitter = \sqrt{\frac{2N}{\pi}}\GaiaRVE \quad .
\end{eqnarray}

For a single star system observed $N$ times, the source radial velocity
jitter represents either the minimum uncertainty with which
\Gaia\ can measure radial velocity for a star of that apparent 
brightness and colour. However for stars with stellar companions,
the radial velocity scatter is a function of orbital parameters 
and observing epochs. Thus we can reasonably assume that the
radial velocity jitter will be a mixture of
two populations: single stars, where the radial velocity jitter
represents the minimum radial velocity noise that \Gaia\ can
measure for a star of that type, and a binary population that
could have radial velocity variances well above the noise floor.

>>>>

\newcommand\teff{T_{\rm eff}}
\newcommand\logg{\log{g}}
\newcommand\feh{[\rm{Fe}/\rm{H}]}
\newcommand\mh{[\rm{M}/\rm{H}]}
\newcommand{\luminosity}{L_\circ}
\newcommand{\radius}{R_\circ}

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\package}[1]{\texttt{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\Gaia}{\project{Gaia}}
\newcommand{\gaia}{\project{gaia}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\rp}{\textsl{rp}}
\newcommand{\bp}{\textsl{bp}}

\newcommand{\NumberOfHigherOrderSystems}{XXX}
\newcommand{\NumberOfSingleLinedSystems}{XXX}

@@

\submitjournal{AAS Journals}

\shorttitle{Stellar multiplicity}
\shortauthors{Casey et al.}

@@

\author{Daniel Foreman-Mackey}
\affiliation{Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

@@

\begin{abstract}
Most stars are in binary systems. The occurrence rate of multiple order systems 
(e.g., binaries, triples) is relevant to many fields of astrophysics, but the 
effects of stellar multiplicity are commonly ignored in the analysis of most 
stellar and galactic surveys. Here we use \Gaia\ data to identify 
\todo{\NumberOfHigherOrderSystems} higher-order (e.g., binary) systems based on
radial velocity or astrometric jitter, photometry, and for a subset of systems, 
we show that the \emph{lack} of reported radial velocities can be a reliable 
indicator of double-lined spectroscopic binary systems. Our sample includes
\todo{\NumberOfSingleLinedSystems} unresolved binary systems detected solely
from radial velocity or astrometric jitter, which have orbital periods up to 
$\approx$3.5\,yrs. We show that the ratio of jitter in radial velocity and
astrometry can be calibrated to estimate inclination angles (and companion masses)
for thousands of systems. \todo{We comment on stellar multiplicity with stellar
metallicity}. \todo{We comment on systems with stellar remnants (e.g. NS/BHs).}
\end{abstract}

@@

There are many ways that binaries can be reliably identified using \Gaia\
data, including radial velocity measurements, photometry, and astrometry.
Here we describe the radial velocity information that forms part of our 
model before describing how photometry and astrometry are incorporated into our
model.

\subsection{SB1 Systems: Single-lined spectroscopic binaries}

Radial velocity measurements are not available for individual epochs in the 
second \Gaia\ data release, but the median and error of radial velocity
measurements are available for 7,224,631 point sources \cite{gaia_rvs}. 
The radial velocity error provided ($\GaiaRVE$; column \texttt{radial\_velocity\_error}) 
is a function of the number of transits $N$ (i.e., the number of observations; 
represented by column \texttt{rv\_nb\_transits}) and the scatter among those $N$ measurements. 
This allows us to calculate the scatter among those measurements, which we will
refer to as the \emph{radial velocity jitter}, for each source
\begin{eqnarray}
\RVJitter = \sqrt{\frac{2N}{\pi}}\GaiaRVE \quad .
\end{eqnarray}

For a single star system observed $N$ times, the source radial velocity
jitter represents either the minimum uncertainty with which \Gaia\ can measure 
radial velocity for a star of that apparent brightness and colour. However for 
stars with stellar companions and no evidence of double lines or binarity in the
cross-correlation-function, so-called SB1 type systems \citep{someone}, the
radial velocity jitter is a function of the orbital parameters and observing
epochs. We assume that the distribution of radial velocity jitter in the second
\Gaia\ data release is a mixture of two populations:

\begin{enumerate}
\item \emph{single-star systems}, where the radial velocity jitter represents the
      minimum radial velocity noise that \Gaia\ can measure for a star of that
      apparent magnitude and colour; and

\item \emph{higher-order star systems}, where the radial velocity jitter is
      significantly above that for single-star systems of a similar apparent
      magnitude and colour.
\end{enumerate}

We make a number of other explicit assumptions about the radial velocity
measurements reported by \Gaia:

\begin{enumerate}
    \item the RV error is or is not a function of colour
    \item the template used for RV determination does or does not affect the 
          error floor
    \item the jitter in radial velocity measurements for a single star will
          increase as a function of apparent flux, etc.
    \item the spread in the jitter will also increase as a function of these
          properties
\end{enumerate}

Given these assumptions, the model we adopt for the radial velocity jitter is
\begin{eqnarray}
    MODEL
\end{eqnarray}

\noindent{}where \todo{X is Y} abd .

In principle for well-sampled circular orbits this model would allow us to
identify binary star systems with orbital periods up to twice the observing
span in the second \Gaia\ data release ($\approx$21 months observing span; 
$\approx$44 month orbital periods). For higher-order star systems with long 
periods, or with mass ratios that result in a small velocity semi-amplitude $K$, 
under our assumptions these systems may be classified as single-star systems 
because the intrinsic radial velocity variation is within the expected 
variations for single-stars of a similar apparent magnitude and colour. We 
revisit this issue in Section \ref{sec:sb_limits}, where we explicitly define 
the probability that a \Gaia\ source is a higher-order star system $p(y|\textrm{SB1})$ 
within defined limits of orbital period (and other orbital parameters). Outside of this
parameter range, we are insensitive to distinguishing single-star systems from
higher-order SB1-type systems.

\begin{itemize}
    \item show jitter compared to RP flux, etc
    \item how the model was fit
    \item show results of the fit
    \item what things did we \emph{not} include in the model (eg difference between temperature and the temperature of the template used), and why
    \item results
\end{itemize}

\subsection{SB2 Systems: Double-lined spectroscopic binaries}

Many sources in the second \Gaia\ data release do not have any radial velocity
measurements reported, despite being bright enough ($G \lesssim 13$) and in a
suitable temperature range for radial velocities to be measured. The principle
reason why radial velocity measurements are \emph{not} reported for these stars
is because the \Gaia/DPAC team have identified the source to be a double-lined
spectroscopic binary (a SB2-type system), in which case it is not sensible to 
report a single radial velocity measurement. For fainter or bluer sources, 
however, the radial velocity may not be reported simply because it is too faint, 
blue, or red.

We calculated the completeness of radial velocity measurements (e.g., the
fraction of sources with reported radial velocities) as a function of all 
available source properties that might affect whether the radial velocity may
be reported or not. This included position ($\alpha$, $\delta$, $l$, $b$),
parallax, proper motions, apparent magnitudes, colours, properties of the radial
velocity templates, stellar parameters ($\teff$, $\radius$, $\luminosity$),
and other properties. The full list of properties investigated is described in
Appendix \todo{X}. In Figure \ref{fig:radial_velocty_completeness} we show the
completeness as a function of just some of these properties, which demonstrate 
that the radial velocity completeness is relatively flat until a star becomes 
too faint (low \rp\ flux), or is either too blue or too red. We adopt 
conservative limits for when the radial velocity completeness starts to drop 
with these properties, and we assume that any source within the following
parameter range

\begin{eqnarray}
    \todo{X} =& \textrm{phot rp mean flux}  = &\todo{X} \\
    \todo{X} =& \textrm{bp}-\textrm{rp}    = &\todo{X}
\end{eqnarray}
\noindent{}is likely to be a double-lined spectroscopic binary (SB2) if no 
radial velocity is reported.

\begin{itemize}
    \item completeness
    \item cross-check against SB catalogues -- does it match expectations?
\end{itemize}

\subsection{Unresolved equal mass binaries: photometry}

\subsection{Astrometric binaries}

Many higher-order star systems are unresolved point sources in \Gaia, but the
stellar companion(s) have sufficient mass to measurably affect the tangential 
velocity of the primary and produce a jitter in astrometric positions. There are
many other sources of astrometric jitter, but here we show that the principle
effect of astrometric excess noise is due to stellar companions. Note that due
to the astrometry `degree of freedom bug' in the processing of the second \Gaia\
data release \citep[e.g., see Appendix A1 of ][]{Lindegren:2018}, we only make
use of astrometric excess noise from the \todo{Tycho-Gaia Astrometric Solution}
\citep{Michalik, TGAS, etc}.

\begin{itemize}
    \item when would we expect AEN? what stellar or orbital properties does it 
          correlate or grow with?
    \item calibration model for AEN in DR1
\end{itemize}
